Abstractions for Type-Level Programming

### 2022-05-12 symposium.tex
Cite Tyre

The Scala standard library contains a package to manipulate regular expressions.
This package is based on Java's implementation, and thus benefits from the high performance of the JVM's regex engine.
The Scala implementation innovates in its presentation: it improves upon Java's solution by providing an ergonomic API based on pattern matching that is both elegant and concise.
The package's documentation starts with the following example:

@@

In this \paper, we propose a new design for Scala's regular expression library which provides a type-safe and null-safe mechanism for capturing group extraction.
Our design makes extensive use of match types\paperOnly{~\citep{blanvillain2022type}, Scala~3's new feature for type-level programming,} to statically analyze regular expressions during type checking.
We build our interface to mimic Scala's original regular expression API so that Scala programmers can use it as a drop-in replacement and enjoy the additional safety with minimal migration costs.

@@

\thesisOnly{
The source code of our implementation is available online\footnote{\texttt{https://github.com/OlivierBlanvillain/regsafe}} under the MIT licence.
}
\paperOnly{
The source code of our implementation is available online\footnote{This URL is hidden for the purpose of the double-blind review process.} under the MIT licence.
}

@@

  \item The runtime capturing group processing component, which extracts and sanitizes the output of the \begin{diff}JVM's\end{diff} regex engine, in accordance with the previously computed type-level representation.
  This component takes the form of a function called |transform|, which we present in \Cref{sec:term-level}.

@@

Rest assured, this is not an oversight!
Regular expressions with trailing backslashes are invalid and result in runtime crashes (reported as an "unexpected internal error", on the JVM).
\begin{diff}
Even though systematic detecting of invalid regular expressions is out of the scope of our library, doing so when possible is still desirable.
\end{diff}
If we invoke |Compile| on a regular expression with a trailing backslash, the compiler will run into a call to |charAt| with an out of bounds argument, catch the corresponding exception and report it as a compilation error, which is certainly better reporting that same error at run-time.
\begin{diff}
At the time of writing, the Scala~3 compiler does not support the customization of match type errors.
% TODO
\end{diff}

@@

% TODO: How understandable are the compiler's error messages if the programmer makes a regex-related mistake? In particular, what's the error message if the programmer provides an incorrect number of arguments to an extractor? What's the error message in case of unbalanced parentheses in the regex? What about other forms of invalid regexes? Could you list the kinds of regex-library-related error messages to be expected?

@@

The Haskell ecosystem contains several packages for regular expressions, with various levels of static safety.
The regex-applicative package stands on the safe side of that spectrum: it allows programmers to write regular expressions using a parser combinator library, which entirely removes the need to reference capturing groups \citep{cheplyaka2011regex}.
The main drawback of this solution lies in the library's learning curve: programmers already familiar with POSIX regular expressions~\citep{ieee2018the} are forced to learn a new syntax, which is arguably more complicated.

@@

% Another type-safe regex library that is not discussed in related work is Tyre: https://github.com/Drup/tyre
% TODO \citep{radanne2020tyre}

>>>>

The Scala standard library contains a package to manipulate regular expressions.
This package is based on Java's implementation, and thus benefits from the high performance of \begin{diff}Java's\end{diff} regex engine.
The Scala implementation innovates in its presentation: it improves upon Java's solution by providing an ergonomic API based on pattern matching that is both elegant and concise.
The package's documentation starts with the following example:

@@

In this \paper, we propose a new design for Scala's regular expression library, which provides a type-safe and null-safe mechanism for capturing group extraction.
Our design makes extensive use of match types\paperOnly{~\citep{blanvillain2022type}, Scala~3's new feature for type-level programming,} to statically analyze regular expressions during type checking.
We build our interface to mimic Scala's original regular expression API, so that Scala programmers can use it as a drop-in replacement and enjoy the additional safety with minimal migration costs.

@@

The source code of our implementation is available online\footnote{\url{https://github.com/OlivierBlanvillain/regsafe}}, under the MIT license.

@@

  \item The runtime capturing group processing component, which extracts and sanitizes the output of \begin{diff}Java's\end{diff} regex engine, in accordance with the previously computed type-level representation.
  This component takes the form of a function called |transform|, which we present in \Cref{sec:term-level}.

@@

Rest assured, this is not an oversight!
Regular expressions with trailing backslashes are invalid and result in runtime crashes (reported as an "unexpected internal error", on the JVM).
\begin{diff}
Even though systematic detecting of invalid regular expressions is out of the scope of our library, doing so when possible is still desirable.
\end{diff}
If we invoke |Compile| on a regular expression with a trailing backslash, the compiler will run into a call to |charAt| with an out of bounds argument, catch the corresponding exception and report it as a compilation error, which is certainly better reporting that same error at run-time.
\begin{diff}
At the time of writing, the Scala~3 compiler does not support the customization of match type errors.
\end{diff}

@@

@@

The Haskell ecosystem contains several packages for regular expressions, with various levels of static safety.
The regex-applicative package stands on the safe side of that spectrum: it allows programmers to write regular expressions using a parser combinator library, which entirely removes the need to reference capturing groups \citep{cheplyaka2011regex}.
The main drawback of this solution lies in the library's learning curve: programmers already familiar with POSIX regular expressions~\citep{ieee2018the} are forced to learn a new syntax, which is arguably more complicated.
\begin{diff}
Tyre~\citep{radanne2020tyre} is an OCaml library that also uses combinators to define type-safe regular expressions.
\end{diff}

@@
